---
title: "Tidying Data in R"
author: "The DSBC team"
output: 
    revealjs::revealjs_presentation:
        slide_level: 1
---

# Tidying data in R
(with a little data wrangling thrown in)

<!--
Notes

# Use the outreach data for now since we need to be careful about making the RCT data public
- check that ggplot2, tidyr and dplyr are installed with Rstudio

- teach the basic use of dplyr
- contrast this with excel?
- use an extract from spotlight?

- Tasks
- =====
- rename a variable
- clean text
    + convert to lower case
    + split and join characters
- concept of factors (categorical data type)
    + ordered 
    + non-ordered
- unnest (from tidyr)
- show case features of dplyr for re-arranging and grouping your data
    + filter
    + summarise by group

 -->

# Set-up (English)

This is just a standard bit of setup. It performs the following tasks.

1. Clean the workspace
2. Load some code libraries 
3. Load some data

# Set-up (code)

```{r setup, eval=TRUE}
rm(list=ls(all=TRUE))       # clean the workspace

require(tidyr)              # Load the tidyr library (for data tidying)
require(dplyr)              # Load the dplyr library (for data wrangling)

# Load the outreach data (stored in a CSV file)
# Need to explain the ../data/ part of the path
ddata <- read.csv("../data/outreach.csv")
```

# Inspect your data


OK. So the data is now in R. You can verify this with the `ls()` command that lists everything in your workspace.

```{r inspect, eval=TRUE}
ls()

# Nice pretty table of your data
tbl_df(ddata)

# Open spreadsheet like view
# View(ddata)

```

# Next

Simple tasks you might wish to do

```{r simple, eval=TRUE}
# require(ggplot2)
# data(diamonds)

# # Sort your data 
# tbl_df(diamonds)
# diamonds %>% arrange(carat) %>% tbl_df()
# diamonds %>% arrange(desc(carat)) %>% tbl_df()

# # Rename a column
# diamonds <- rename(diamonds, length=x)
# diamonds <- rename(diamonds, width=y)
# diamonds <- rename(diamonds, depth=z)
# tbl_df(diamonds)

# # Filter your data (subset observations)
# # will need to teach <,>,==,<=,>=
# filter(diamonds, carat>4)

# # Subset columns
# data(iris)
# tbl_df(iris)
# select(iris, ends_with("width"))
# head(select(iris, ends_with("width")))

# # Summarise data
# summarise(diamonds, avg=mean(carat))
# count(diamonds,cut)
# summarise_each(diamonds, funs( min, max ))

# diamonds %>% group_by(color) %>% summarise(avg = mean(price))

```

# Next

Things which are difficult to do in Excel but very useful

```{r tidyr, eval=TRUE}
library(tidyr)
```

```{r cell_clean, eval=TRUE}
# tbl_df(diamonds)
# d2 <- unite(diamonds, cut_color, c(cut, color))
# tbl_df(d2)
# d1 <- separate(d2, cut_color, into=c(cut, col), sep="_")
# head(d1)

```




